---
title: "Diabetes dataset - SLR"
author: "Haley Egan"
date: "12/5/2021"
output: html_document
---
```{r}
#clean data 

library(faraway)
library(caret)

df <- diabetes

?diabetes
summary(df)
head(df)

df <- subset(df, select = -c(id,bp.2s,bp.2d))

df <- df                                    
df$glyhb[is.na(df$glyhb)] <- median(df$glyhb, na.rm = TRUE)  

df <- df                                    
df$chol[is.na(df$chol)] <- median(df$chol, na.rm = TRUE)  

df <- df                                    
df$hdl[is.na(df$hdl)] <- median(df$hdl, na.rm = TRUE)  

df <- df                                    
df$ratio[is.na(df$ratio)] <- median(df$ratio, na.rm = TRUE)  

df <- df                                    
df$height[is.na(df$height)] <- median(df$height, na.rm = TRUE)  

df <- df                                    
df$weight[is.na(df$weight)] <- median(df$weight, na.rm = TRUE)  

df <- df                                    
df$bp.1s[is.na(df$bp.1s)] <- median(df$bp.1s, na.rm = TRUE)  

df <- df                                    
df$bp.1d[is.na(df$bp.1d)] <- median(df$bp.1d, na.rm = TRUE)  

df <- df                                    
df$waist[is.na(df$waist)] <- median(df$waist, na.rm = TRUE)  

df <- df                                    
df$hip[is.na(df$hip)] <- median(df$hip, na.rm = TRUE)  

df <- df                                    
df$time.ppn[is.na(df$time.ppn)] <- median(df$time.ppn, na.rm = TRUE)  

#df$frame <- df$frame %>% replace_na('medium')

head(df)
#summary(df)
```


#select two columns for Simple Linear Regression
```{r}
#make a new data frame with only glyhb and stab.glu
SLRdf <- df[,c('stab.glu', 'glyhb')]
head(SLRdf)
```

Response Variable = glyhb (Glycosolated Hemoglobin)
Predictor Variable = stab.glu (Stabilized Glucose)

```{r}
#scatter plot to visualize variables with fitted regression line 
ggplot(SLRdf, aes(x=stab.glu,y=glyhb))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+ #fit regression equation line to plot 
  labs(x="Stabilized Glucose", y="Glycosolated Hemoglobin", title="Scatterplot of Glycosolated Hemoglobin against Stabilized Glucose")
```

#Observation:

There appears to be a positive linear relationship between the variables. However, there is clustering in the bottom left of the scatter plot. 


```{r}
##Fit a simple linear regression model
result<-lm(glyhb~stab.glu, data=SLRdf)
summary(result)
```

Estimated Regression Equation: y = 2.2351 + 0.0312x

#Interpretation:

For every unit increase in Stabilized Glucose, Glycosolated Hemoglobin will increase by 2.2351. There is a positive linear relationship between the two variables.


# Model diagnostics 

```{r}
##store fitted y & residuals
yhat<-result$fitted.values
res<-result$residuals

##add to data frame
SLRdf<-data.frame(SLRdf,yhat,res)
```

```{r}
##residual plot
ggplot(SLRdf, aes(x=yhat,y=res))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  labs(x="Fitted y", y="Residuals", title="Residual Plot")
```

#Interpretation:

The error terms are clustered on the left side of the plot. This could indicate that the error terms do not have a constant variance, and that assumption 3 is not met. There does appear to be a linear pattern (seen in scatter plot), but there is clustering on the left side, which does not meet assumption 1, that the residuals are evenly scattered around the line without an apparent pattern. Also, the error terms appear to have a mean of zero, which does meet assumption 2. A transformation is needed to meet the assumptions.

```{r}
#box cox plot 
library(MASS) ##to use boxcox function
boxcox(result)
```

#Interpretation: 

Zero does not fall within the box cox plot lines, so a log transformation is not necessary. However, a transformation is needed. -1 will be used for the transformation since it is within the lines. 

```{r}
##transform y and then regress ystar on x
ystar<-(SLRdf$glyhb)^(-0.8)
SLRdf<-data.frame(SLRdf,ystar)
result.ystar<-lm(ystar~stab.glu, data=SLRdf)

##store fitted y & residuals
yhat2<-result.ystar$fitted.values
res2<-result.ystar$residuals

##add to data frame
SLRdf<-data.frame(SLRdf,yhat2,res2)
```

```{r}
##Fit a regression model
result2<-lm(ystar~res2, data=SLRdf)

##store fitted y & residuals
yhat<-result2$fitted.values
res<-result2$residuals

##add to data frame
SLRdf<-data.frame(SLRdf,yhat,res)

##residual plot
ggplot(SLRdf, aes(x=yhat,y=res))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  labs(x="y*", y="Residuals", title="Y* Residual Plot")
```

```{r}
boxcox(result.ystar)
```

#Interpretation:

The transformation did improve the residual plot. However, there is still clustering on the left side of the plot. Further transformation is needed.


```{r}
##transform x and then regress y* on x*
xstar<-sqrt((SLRdf$stab.glu)^1)
SLRdf<-data.frame(SLRdf,xstar)
result2.xstar<-lm(ystar~stab.glu, data=SLRdf)

##store fitted y & residuals
xhat2<-result2.xstar$fitted.values
res2<-result2.xstar$residuals

##add to data frame
SLRdf<-data.frame(SLRdf,yhat2,res2)
```
#Create a scatterplot of y∗ against x*. 
```{r}
##scatterplot, and overlay regression line
ggplot(SLRdf, aes(x=xstar, y=ystar))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+ #fit regression equation line to plot 
  labs(x="x*", y="y*", title="Scatterplot of y* against x*")
```

```{r}
##residual plot with ystar
ggplot(SLRdf, aes(x=xstar,y=res2))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  labs(x="Fitted y", y="Residuals", title="X* Residual Plot")
```


#Interpretation:

The transformation on x and y improved the residual plots. However, there is still some clustering. 


#Check other assumptions

```{r}
#ACF plot of residuals 
acf(res2, main="ACF Plot of Residuals with ystar")
```

#Interpretation:

The ACF plot checks assumption 4, that the error terms are uncorrelated and independed. The assumption is nearly met, because the vertical lines are within the horizontal blue lines. However, one line is not within the blue zone.

```{r}
#QQ plot of residuals
qqnorm(res2)
qqline(res2, col="red")
```

#Interpretation:

The QQ plot checks assumption 5, that the errors follow normal distribution. A normal probability plot should fall close to the red line, representing expected values under normality. The errors almost entirely fall on the red line, except on the far right side where the errors curve upward, away from the line. The assumption is close to being met. 


```{r}
result.ystar
```


```{r}
#correlation b/w glycosolated hemoglobin and stabilized glucose
cor(SLRdf$glyhb, SLRdf$stab.glu)
```

#Interpretation:

There is a strong positive relationship between Glycosolated Hemoglobin and Stabilized Glucose, because 0.749824 is fairly close to 1. 

```{r}
#average of residuals
mean(result$residuals)
```

#Interpretation:

The average value of all the residuals is 1.399475e-17. This suggests that the average data point is above, but very close the fitted line, because 1.399475e-17 is very close to 0. On average, the line appears to fit the data well. 

```{r}
#ANOVA table

anova.tab<-anova(result)
anova.tab

#calculate SST
SST<-sum(anova.tab$"Sum Sq")
SST

#calculate R2
anova.tab$"Sum Sq"[1]/SST
```


#Hypothesis Test:
H0: B1 = 0;
Ha: B1 ≠ 0

- H0, the null hypothesis, suggests that there is no evidence of a linear relationship between Glycosolated Hemoglobin and Stabilized Glucose. (Regression line would be flat).
- Ha, the alternative hypothesis, suggests that there is evidence of a linear relationship between Glycosolated Hemoglobin and Stabilized Glucose. (Regression line would not be flat).

p-value = 2.2e-16 

Based on the Anova F test for this SLR, we reject the null hypothesis and conclude that there is a linear relationship between Glycosolated Hemoglobin and Stabilized Glucose. The p-value is very low, at 2.2e-16, less than any alpha, so it is unlikely that the slope of 0.031221 occurred by chance. 


```{r}
##produce 95% CIs for all regression coefficients
confint(result,level = 0.95)
```

#Interpretation:
There is 95% confidence that the Stabilized Glucose level will be between (0.02851678, 0.03392596).

```{r}
#confidence interval for average glycosolated hemoglobin with a stabilized glucose level of 0.04
##to produce 95% CI for the mean response when x=0.04
##and the 95% PI for the response of an observation when x=0.04
newdata<-data.frame(stab.glu=0.04)
predict(result,newdata,level=0.95, interval="confidence") #newdata <- for what value of x do i want to make a prediction of y for 

#For a prediction interval for average glycosolated hemoglobin with a stabilized glucose level of 0.04
predict(result,newdata,level=0.95, interval="prediction") 

```

The 95% confidence interval for average Glycosolated Hemoglobin with a Stabilized Glucose level of 0.04 is (1.914303 2.558441).

The prediction interval for average  Glycosolated Hemoglobin with a Stabilized Glucose level of 0.04 with 95% confidence is (-0.6597715, 5.132515).
























